Secure Composition \\ of Robust and Optimising Compilers

### 2024-05-28 csf2/main.tex
address some todos

Future work intends to use upper and lower composition as means to build a secure compiler incrementally.
\MPin{
	we should add briefly which real-world compilers (taken from the elided examples) are upper and lower compos
}
\MP{no coq here ?}

@@

\gls*{ss}~\cite{patrignani2021exorcising}, similar to \gls*{scct}, is an overapproximation of a variant of noninterference.
The definition of the property itself turns out to be syntactically equal to \gls*{scct}, modulo the underlying trace model.
\MPin{recap its meaning for security}

@@

  \begin{nscenter}
  $
    \sssafe:=\left\{\trace[_{\mathghost}] \left|\begin{array}{rcl}
        \trace[_{\mathghost}]=\hole{\cdot} & \text{ or } & \trace[_{\mathghost}]=\preevent[_{\mathghost}];\unlock\cdot\trace[_{\mathghost}'] \wedge \trace[_{\mathghost}']\in\sssafe
                                 \end{array}\right.\right\}
  $ 
  \end{nscenter}
\end{definition}
\MPin{why isn't it ok to have locks with NONE in the trace?}

@@

All presented programming languages share a common fragment which is partially presented here and in full detail in the technical report. 
The trace models of all languages are similar to those presented earlier (\Cref{sec:compprop}).
One technical detail is the addition of a sandboxtag, indicating who is to blame for an emitted action: context ($\ctx$) or component ($\comp$).
This is a standard technique in secure compilation in order to rule out irrelevant context-level events. 
For example, a context immediately deallocating an allocated memory region twice trivially violates memory safety, but the main interest in secure compilation is at component-level events and if those signal, e.g., a memory safety issue.

\MPin{this section does not define anything, not even the sandbox tag metavariable}
\MPin{
	move b/t/a/e/F/CHI/H/Delta/Phi/Psi to here.
	their explanation too.
	make black.
}
\MPin{
	what are pointer maps and Control states?
}

@@

$\src{L_{\tmssafe}}$ is a relatively naive language with second class functions, the only values being numbers. 
It has simple binary operations ($\src{\lbinop{\expr[_1]}{\expr[_2]}}$), branches ($\src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}}$), C-like function calls ($\src{\lcall{f}{\expr}}$) and returns ($\src{\lreturn{\expr}}$), as well as C-like memory operations. 
%Pointers are left second-class and can only be referenced by an identifier, so addresses cannot be guessed by contexts.
Programs have sets of pre-determined functions called libraries and it is assumed that functions are marked as being part of some component ($\src{\library_{\comp}}$) or context ($\src{\library_{\ctx}}$).
Functions ($(\src{x};\src{\natt\to\natt};\src{\expr})$) are modelled as tuples and can only accept integers, which also means that contexts may not pass pointers to the component.
%Extending this is orthogonal to this work.

\vspace{-.5em}
{
% \small
\[
  \begin{array}{rrcl}
    \text{(Base-Events)} & \src{\preevent} &:=& \src{\ev{Alloc \loc\ n}} \mid \src{\ev{Dealloc \loc}} \mid \src{\ev{Get \loc\ n}} \mid \src{\ev{Set \loc\ n}} \\
    \text{(Control Tags)} & \src{\sandboxtag} &:=& \src{\ctx} \mid \src{\comp} \\
    \text{(Events)} & \src{\event} &:=& \src{\preevent;\sandboxtag} \mid \src{\emptyevent} \mid \src{\lightning} \\ 
    \text{(Values)} & \src{v} &:=& \src{n} \\
    \text{(Expressions)} & \src{\expr} &:=& \src{x} \mid \src{n} \mid \src{\lbinop{\expr[_1]}{\expr[_2]}} \mid \src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}} \\ 
                         &&&\mid \src{\llet{x}{\expr[_1]}{\expr[_2]}} \mid \src{\llet{x}{new\ \expr[_1]}{\expr[_2]}} \\
                         &&&\mid \src{\ldelete{x}} \mid \src{\lget{x}{\expr}} \mid \src{\lset{x}{\expr[_1]}{\expr[]}} \mid \src{stuck}\\
                         &&&\mid \src{\lcall{f}{\expr}} \mid \src{\lreturn{\expr}} \\
    \text{(Functions)} & \src{F} &:=& (\src{x};\src{\natt\to\natt};\src{\expr}) \\
    \text{(Libraries)} &&& \src{\library_{\ctx}} : \src{Vars} \to \src{F} \hspace{2em}
    \src{\library_{\comp}} : \src{Vars} \to \src{F} \\
    \text{(Heaps)} & \src{H} &:=& \src{\hole{\cdot}} \mid \src{n},\src{H}\hspace{0.5cm}    \text{(Pointer Maps)}\ \ \ \src{\Delta} \\
    \text{(Memory States)} & \src{\memstate} &:=& \left(\src{H^{\ctx}};\src{H^{\comp}};\src{\Delta}\right)\\
  \text{(Control States)} & \src{\cfstate} & & \text{(States)}\ \ \ \src{\Omega}\ :=\ \left(\src{\cfstate};\src{\sandboxtag};\src{\memstate}\right)\\
  \end{array}
\]
}

Crucially, $\src{L_{\tmssafe}}$ is statically typed, where the type system is inspired by $L^{3}$~\cite{morrisett2005L3,scherer2018fabulous}.
The type system of $\src{L_{\tmssafe}}$ enforces that every well-typed $\src{L_{\tmssafe}}$ program satisfies \gls*{tms} (\Cref{thm:wt:tms}).

@@

Operational state ($\src{\Omega}$) is a triple consisting of control-flow state ($\src{\cfstate}$), a control tag ($\src{\sandboxtag}$), and a memory state ($\src{\memstate}$), which carries information about the memory.
Because pointers are kept ``alive'' in pointer maps ($\src{\Delta}$), $\Ltms$ dynamic semantics does not get stuck if it encounters a memory-safety violation such as a double--free.
The memory state carries two heaps to model sandboxing between a context ($\src{H^\ctx}$) and a component ($\src{H^\comp}$).

The trace model is largely similar to the memory safety trace model (\Cref{subsec:basic:memsafety:tracemodel}), where the only real differences are separate events for reading and writing memory, instead of just using it, and the sandboxtags.
With these tags, it is possible to ignore effects performed by a spurious context that, e.g., immediately deallocates a memory location twice, thus violating \gls*{tms}.
Even though this could be used for blame preservation~\cite{patrignani2023blame}, this beyond the focus of this paper.

@@

\MPin{
	can this be made formal?
	what's the point?
}

@@

With the advent of speculative execution in processors, it is possible to exploit this unusual paradigm leak secrets.
This work has taken inspiration in the sematics of $\Lspec$ and in \Cref{def:trace:ss} from recent formalisations~\cite{guarnieri2018spectector,fabian2022automatic}.
\MPin{
  we take SS from exorcising.
  we do not take SNI from spectector because it is a hypersafety.
  other work defined similar props.
}
% A subsequent line of work~\cite{fabian2022automatic} shows how to modularly define and compose formal semantics that models different speculative execution variants. 
% Recent work~\cite{fabian2024lift} demonstrates how to lift existing secure compilation proofs for security properties related to speculative execution to more powerful properties, i.e., properties defined for languages with a trace model that enables more vulnerabilities, such as combinations of different speculative execution exploits. 
% To this end, they define a lifting framework and demonstrate how a secure compilation proof for a weaker property can be lifted to a stronger one. 
% Contrary to the compositionality framework presented in this paper, these works are only concerned with composing semantics of the respective programming languages, while in this paper, the secure compilers themselves are composed.

>>>>

Future work intends to use upper and lower composition as means to build a secure compiler incrementally
This follows current non-formal-methods based compiler engineering, such as LLVM~\cite{lattner2004llvm}, where languages such as \texttt{C++} and \texttt{Fortran} translate into LLVM IR and languages such as \texttt{x86} or \texttt{ARMv8} assembly variants translate out of LLVM IR.

% no coq, because above definitions are non-formal. 
%   How would you precisely define the lower compo compiler, given two compilers?

@@

\gls*{ss}~\cite{patrignani2021exorcising}, similar to \gls*{scct}, is a sound overapproximation of a variant of noninterference.

@@

  \begin{nscenter}
  $
    \sssafe:=\left\{\trace[_{\mathghost}] \left|\begin{array}{l}
      \trace[_{\mathghost}]=\hole{\cdot} \text{ or } \exists\trace[_{\mathghost}'].\\
      \left(\trace[_{\mathghost}]=\preevent[_{\mathghost}];\unlock\cdot\trace[_{\mathghost}'] \text{or }\trace[_{\mathghost}]=\preevent[_{\mathghost}];\lock_{\text{NONE}}\cdot\trace[_{\mathghost}']\right)\\
      \text{and }\ \trace[_{\mathghost}']\in\sssafe
                                 \end{array}\right.\right\}
  $ 
  \end{nscenter}
\end{definition}
The technical setup so far leads to the above definition, where only locks annotated with $\text{SPECTRE-PHT}$ are disallowed to occur on the trace.
That way, programs attaining \gls*{ss} do not necessarily attain \gls*{scct}.

@@

All presented programming languages share a common fragment which is partially presented here and in full detail in the technical report.

\vspace{-.5em}
{
  \renewcommand{\src}[1]{\mathsf{#1}}
% \small
\[
  \begin{array}{rrcl}
    \text{(Base-Events)} & \src{\preevent} &:=& \src{\ev{Alloc \loc\ n}} \mid \src{\ev{Dealloc \loc}} \mid \src{\ev{Get \loc\ n}} \mid \src{\ev{Set \loc\ n}} \\
    \text{(Control Tags)} & \src{\sandboxtag} &:=& \src{\ctx} \mid \src{\comp} \\
    \text{(Events)} & \src{\event} &:=& \src{\preevent;\sandboxtag} \mid \src{\emptyevent} \mid \src{\lightning} \\ 
    \text{(Values)} & \src{v} &:=& \src{n} \\
    \text{(Expressions)} & \src{\expr} &:=& \src{x} \mid \src{n} \mid \src{\lbinop{\expr[_1]}{\expr[_2]}} \mid \src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}} \\ 
                         &&&\mid \src{\llet{x}{\expr[_1]}{\expr[_2]}} \mid \src{\llet{x}{new\ \expr[_1]}{\expr[_2]}} \\
                         &&&\mid \src{\ldelete{x}} \mid \src{\lget{x}{\expr}} \mid \src{\lset{x}{\expr[_1]}{\expr[]}} \mid \src{stuck}\\
                         &&&\mid \src{\lcall{f}{\expr}} \mid \src{\lreturn{\expr}} \\
    \text{(Functions)} & \src{F} &:=& (\src{x};\src{\expr}) \\
    \text{(Libraries)} &&& \src{\library_{\ctx}} : \src{Vars} \to \src{F} \hspace{2em}
    \src{\library_{\comp}} : \src{Vars} \to \src{F} \\
    \text{(Heaps)} & \src{H} &:=& \src{\hole{\cdot}} \mid \src{n},\src{H}\hspace{0.5cm}    \text{(Pointer Maps)}\ \ \ \src{\Delta} \\
    \text{(Memory States)} & \src{\memstate} &:=& \left(\src{H^{\ctx}};\src{H^{\comp}};\src{\Delta}\right)\\
  \text{(Control States)} & \src{\cfstate} & & \text{(States)}\ \ \ \src{\Omega}\ :=\ \left(\src{\cfstate};\src{\sandboxtag};\src{\memstate}\right)\\
  \end{array}
\]
}
{
  \renewcommand{\src}[1]{\mathsf{#1}}
The trace models of all languages are similar to those presented earlier (\Cref{sec:compprop}).
One technical detail is the addition of a control tag, indicating who is to blame for an emitted action: context ($\ctx$) or component ($\comp$).
This is a standard technique in secure compilation in order to rule out irrelevant context-level events. 
For example, a context immediately deallocating an allocated memory region twice trivially violates memory safety, but the main interest in secure compilation is at component-level events and if those signal, e.g., a memory safety issue.
Even though this tagging could be used for blame preservation~\cite{patrignani2023blame}, this beyond the focus of this paper.
Another key difference is that memory accesses are now explicitly modeled as reads ($\src{\ev{Read\ \loc\ n}}$) and writes ($\src{\ev{Write\ \loc\ n}}$) instead of just uses.

All languages have at least numbers as values ($\src{v}$) and second class functions ($\src{F}$).
Functions are modeled as pairs containing the name of one argument and the body of the function.
Bodies are just ordinary expressions, which can be simple binary operations ($\src{\lbinop{\expr[_1]}{\expr[_2]}}$), conditionals ($\src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}}$), C-like function calls ($\src{\lcall{f}{e}}$) and returns ($\src{\lreturn{\expr}}$), as well as C-like memory operations. 
Programs have sets of pre-determined functions called libraries and it is assumed that functions are marked as being part of some component ($\src{\library_{\comp}}$) or context ($\src{\library_{\ctx}}$).
Moreover, operational state ($\src{\Omega}$) is a triple consisting of a control-flow state ($\src{\cfstate}$), a control tag ($\src{\sandboxtag}$), and a memory state ($\src{\memstate}$). 
The latter carrying information about pointers that are kept ,,alive'' in pointer maps ($\src{\Delta}$), so that dynamic semantics do not get stuck when encountering, e.g., a double-free.
The memory state also carries two heaps to model sandboxing between a context ($\src{H^{\ctx}}$) and a component ($\src{H^{\comp}}$).

}

@@

$\src{L_{\tmssafe}}$ is the only statically typed language in this case study and restricts functions ($\src{F}$) to the typing signature $\src{\natt\to\natt}$. 
The type system of $\src{L_{\tmssafe}}$ is inspired by $L^{3}$~\cite{morrisett2005L3,scherer2018fabulous} and enforces that every well-typed $\src{L_{\tmssafe}}$ program satisfies \gls*{tms} (\Cref{thm:wt:tms}).

@@

@@

%\MPin{
%	can this be made formal?
%	what's the point?
%}

@@

With the advent of speculative execution in processors, it is possible to exploit this unusual paradigm leak secrets.
This work has taken inspiration in the sematics of $\Lspec$ and in \Cref{def:trace:ss} from recent formalisations~\cite{guarnieri2018spectector,fabian2022automatic}.
% \MPin{
%  we take SS from exorcising.
%  we do not take SNI from spectector because it is a hypersafety.
%  other work defined similar props. 
% }
% A subsequent line of work~\cite{fabian2022automatic} shows how to modularly define and compose formal semantics that models different speculative execution variants. 
% Recent work~\cite{fabian2024lift} demonstrates how to lift existing secure compilation proofs for security properties related to speculative execution to more powerful properties, i.e., properties defined for languages with a trace model that enables more vulnerabilities, such as combinations of different speculative execution exploits. 
% To this end, they define a lifting framework and demonstrate how a secure compilation proof for a weaker property can be lifted to a stronger one. 
% Contrary to the compositionality framework presented in this paper, these works are only concerned with composing semantics of the respective programming languages, while in this paper, the secure compilers themselves are composed.
